Topological order in the quantum dimer-pentamer model

### 2016-04-20 qdpm_paper.tex
Worked on background section

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

@@

\begin{figure}[t!]
    \centering
    \includegraphics[width=0.5\columnwidth]{QDPM_ex_config.pdf}
    \caption{Example of an allowed configuration in the quantum dimer-pentamer model.}
    \label{fig:QDPMex}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background}
\label{sec:Background}

\subsection{Generalized quantum dimer models}
\label{sec:GQDM}

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.8\linewidth]{example_local_constraints.pdf}
    \caption{Clockwise from the upper left panel we show an example of a QDM with constraints: one
    dimer ($n_0 = 1$), two dimers ($n_0 = 2$), and an even ($n_P = 0$) or odd ($n_P = 1$) number of dimers touching each vertex.}
    \label{fig:example_local_constraints}
\end{figure}

Here we will provide background about generalized quantum dimer models on the square lattice. We consider systems with Ising-like dimer degrees of freedom on each link, such that each link has either $0$ or $1$ dimers. The Hilbert space consists of configurations of dimers on the square lattice that satisfy a local constraint at each vertex. Defining $n_v$ as the number of dimers that touch a vertex, we can distinguish between fixed number constraints of the form $n_v = n_0$ and fixed parity constraints of the form $n_v\mod 2 = n_P$. The unique, non-trivial constraints of this form for the square lattice are the number constraints $n_0=1,2$ and the parity constraints $n_P =0,1$ (see \figref{fig:example_local_constraints}).

@@

Because of the local constraints, all states in the Hilbert space of a GQDM are invariant under local gauge transformations a the vertices, $G_v$~\cite{Moessner2001}. For GQDM with a number constraint, we write the gauge transformation as:
\begin{equation}
G_v = e^{i \alpha \left( n_v -n_0 \right)}; \notag
\end{equation}
as the states in the GQDM Hilbert space are invariant under $G_v$ for all $0 \leq \alpha \leq 2 \pi$, this represents a local $U(1)$ gauge symmetry. When the number constraint is relaxed to a parity constraint, there is a correspondingly reduced gauge symmetry which can be written as
\begin{equation}
G_v^P = e^{\pm i \pi \left(n_v-n_P\right)}, \notag
\end{equation}
and the gauge symmetry is now reduced to to $Z_2$.
By construction, the Hamiltonian is gauge invariant and thus commutes with $G_v$
\begin{equation}
\left[ H_{\rm{GQDM}}, G_v \right] = 0. \notag
\end{equation}

Given the gauge symmetry, one may construct a gauge invariant string operator along an oriented path $\Gamma$ path on the dual lattice
\begin{equation}
S_\Gamma \equiv \prod_{\ell \in \Gamma} e^{i \alpha_\ell n_\ell}
\end{equation}
where $\alpha_\ell$ is a phase that is associated with each link on the lattice, as well as the direction the link is crossed in $\Gamma$.

\subsection{Square lattice QDM}

the simplest constraint is that of the hard-core fully-packed QDM where a single dimer must touch each vertex. We write this constraint as $n_v = 1$,

For example in the QDM $\hat{T}$ is the sum of all possible ``plaquette flip" terms acting on plaquettes with parallel dimers:
\begin{equation}
    \label{}
    \hat{T}_{\rm{QDM}} = \sum_{\bplqt} 
            \left|
                \vrtplqt
            \right\rangle
            \left\langle
                \hrzplqt
            \right|
            +
            h.c..
\end{equation}

>>>>

\newcommand{\HGQDM}{H_\mathrm{GQDM}}
\newcommand{\HQDM}{H_\mathrm{QDM}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t!]
    \centering
    \includegraphics[width=0.5\columnwidth]{QDPM_ex_config.pdf}
    \caption{Example of an allowed configuration in the quantum dimer-pentamer model.}
    \label{fig:QDPMex}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background}
\label{sec:Background}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Generalized quantum dimer models}
\label{sec:GQDM}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.8\linewidth]{example_local_constraints.pdf}
    \caption{Clockwise from the upper left panel we show an example of a QDM with constraints: one
    dimer ($n_0 = 1$), two dimers ($n_0 = 2$), and an even ($n_P = 0$) or odd ($n_P = 1$) number of dimers touching each vertex.}
    \label{fig:example_local_constraints}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Here we will provide background about generalized quantum dimer models on the square lattice. We consider systems with Ising-like dimer degrees of freedom on each link, such that each link has either $0$ or $1$ dimers. The Hilbert space consists of configurations of dimers on the square lattice that satisfy a local constraint at each vertex. Defining $n_v$ as the number of dimers that touch a vertex, we can distinguish between fixed number constraints of the form $n_v = n_0$ and fixed parity constraints of the form $n_v\mod 2 = n_P$ (see \figref{fig:example_local_constraints}).

@@

Because of the local constraints, all states in the Hilbert space of a GQDM are invariant under local gauge transformations a the vertices, $G_v$~\cite{Moessner2001} of the general form
\begin{equation}
G_v = e^{i \alpha \left( n_v -c_0 \right)}; \notag
\end{equation}
where the $c_0$ and $\alpha$ depend on the constraint.  
By construction, the Hamiltonian is gauge invariant and thus commutes with $G_v$
\begin{equation}
\left[ H_{\rm{GQDM}}, G_v \right] = 0. \notag
\end{equation}

The gauge symmetry allows one to construct a gauge invariant string operator $S_\gamma$ along an oriented path $\gamma$ path on the dual lattice. We first orient all the links of the square lattice such that they point from the $A$ sublattice to the $B$ sublattice (see \figref{fig:orientedSL}) and define an oriented dimer flux $\Phi_\gamma$ through the path:
\begin{equation}
\Phi_\gamma \equiv \sum_{\ell \in \gamma} \rm{sgn}\left(\ell,\gamma\right) n_\ell
\end{equation}
where $\mathrm{sgn}(\ell,\gamma)$ is positive (negative) if $\ell$ is oriented to the right with respect to the orientation of $\gamma$. The string operator is defined as
\begin{equation}
S_\gamma \equiv e^{i \alpha \Phi_\gamma}
\end{equation}
where $\alpha$ such that $S_\gamma$ commutes with $\HGQDM$ everywhere except at the ends of $\gamma$. Therefore, for closed loops $\Gamma$, $[ S_\Gamma,\HGQDM]=0$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.8\linewidth]{QDPM_ex_config.pdf}
    \caption{\textcolor{red}{Replace with a figure showing the orientations of the lattice links, with an open oriented string on the dual lattice as well as a closed oriented loop and topologically non-trivial loops.}The link orientations of the square lattice.}
    \label{fig:orientedSL}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Fixed number constraints}

The two unique fixed-number constraints on the square lattice $n_0 = 1$ and $n_0=2$ lead to the QDM and the quantum full-packed loop model (QFPLM), respectively. For such GQDMs with a number constraint, we write the gauge transformation as:
\begin{equation}
G_v = e^{i \alpha \left( n_v -n_0 \right)}; \notag
\end{equation}
as the states in the GQDM Hilbert space are invariant under $G_v$ for all $0 \leq \alpha \leq 2 \pi$, this represents a local $U(1)$ gauge symmetry. The Hamiltonian for these models gives dynamics to parallel dimers on the plaquettes of the lattice:
\begin{align}
    \label{}
   \HQDM = &-t\sum_{\bplqt} \biggl( \eket{\vrtplqt} \ebra{\hrzplqt} +h.c. \biggr) \notag\\
    					&+v\sum_{\bplqt} \biggl( \eket{\vrtplqt} \ebra{\vrtplqt} +\eket{\hrzplqt} \ebra{\hrzplqt} \biggr).
\end{align}
In this case of a fixed number constraint, the Hamiltonian conserves the dimer flux through a closed loop:
\begin{equation}
\left[ \Phi_\Gamma, \HQDM \right] = 0
\end{equation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Fixed parity constraints}

When the number constraint is relaxed to a parity constraint, there is a correspondingly reduced gauge symmetry which can be written as
\begin{equation}
G_v^P = e^{\pm i \pi \left(n_v-n_P\right)}, \notag
\end{equation}
and the gauge symmetry is now reduced to to $Z_2$.
